---
title: "Practical B"
date: "Datatypes and syntax"
output: html_document
---

---

#### Exercises
In this practical we are going to play around with the different types of elements in R. 

---

 1. **Make two vectors: one named `vec1` with values 1 through 6 and one named `vec2` with letters A through F. **
```{r}
vec1 <- c(1, 2, 3, 4, 5, 6)
vec2 <- c("A", "B", "C", "D", "E", "F")
```

To create a vector we used `c()`, which stands for 'concatenation'. It is just a series of numbers or letters. 

---

 2. **Create two matrices, one from `vec1` and one from `vec2`. The dimensions for both matrices are 3 rows by 2 columns. **

```{r}
mat1 <- matrix(vec1, nrow = 3, ncol = 2)
mat2 <- matrix(vec2, nrow = 3, ncol = 2)
```

To create a matrix we used `matrix()`. For a matrix we need to specify the dimensions (in this case 3 rows and 2 columns) and the input (in this case `vec1` or `vec2`) needs to match these dimensions. 

---

3. **Inspect your vectors and matrices. Are all numerical?**

```{r}
vec1
vec2
mat1
mat2
```

`vec1` and `mat1` contain numbers and `vec2` and `mat2` contain characters. 

---

 4. **Make a matrix from both `vec1` and `vec2` with 6 rows and 2 columns. Inspect this matrix.**

```{r}
mat3 <- matrix(c(vec1, vec2), 6, 2)
is.matrix(mat3)
mat3
```
or
```{r}
mat3b <- cbind(vec1, vec2)
is.matrix(mat3b)
mat3b
```


If one or more elements in the matrix represent characters, all other elements are also converted to characters. A matrix is just for either numeric or character elements. Notice that the second approach (the column bind approach from `mat3b`) returns a matrix where the column names are already set to the name of the bound objects. 

To solve the problem of charactered numbers we can create a dataframe. A dataframe is essentially a matrix that allows for character elements. The use of a dataframe is often preferred over the use of a matrix in `R`, except for purposes where pure numerical calculations are done, such as in matrix algebra. However, most datasets do contain character information and a dataframe would normally be your preferred choice when working with your own collected datasets in R. 

---

5. **Make a dataframe called `dat3` where `vec1` and `vec2` are both columns. Name the columns `V1` and `V2`, respectively. Use function `data.frame()`.**

```{r}
dat3 <- data.frame(V1 = vec1, V2 = vec2)
dat3
```

---

6. **Again, make a dataframe called `dat3b` where `vec1` and `vec2` are both columns. Name the columns `V1` and `V2`, respectively. Use function `as.data.frame()` on the matrix obtained from `Question 4`. **

This is a tricky situation. At face value, everything may seem to be in order. But, be aware that the code
```{r}
dat3b <- as.data.frame(mat3b)
dat3b
names(dat3b) <- c("V1", "V2")
dat3b
```
does not work properly (at least not as intended) as the matrix nature of `mat3` turned everything into a character value and you have lost the numerical nature of `vec1`. It may appear to be working, but if we check if column 1 is numerical, it turns out not to be the case. 

---

7. **Check if the first column in the data frames from Question 5 and Question 6 are indeed numeric. If not, determine what they are. **

```{r}
is.numeric(dat3[, 1])
is.numeric(dat3b[, 1])
```
The first column in matrix `dat3b` obtained from Question 5 is indeed not numeric. Instead of asking "Is object *X* of type *Y*?" we simply ask `R` to report the type of the object of interest:
```{r}
typeof(dat3b[,1])
```

Previous versions of these practicals had to deal with factors at this point, since there was an implicit `stringsAsFactors=TRUE` in for example the `as.matrix` function. For good reasons, the behavior of `R` in this regards has changed. (If you are interested, see the following [blogpost](https://blog.r-project.org/2020/02/16/stringsasfactors/) by Kurt Hornik.)

---

8. **Select 1) the third row, 2) the second column and 3) the intersection of these two in the dataframe `dat3` that you have created in Question 5. **
```{r}
dat3[3, ] #3rd row
dat3[, 2] #2nd column
dat3$V2   #also 2nd column
dat3[3,2] #intersection
```

The `[3,2]` index is very useful. The first number (before the comma) represents the row and the second number (after the comma) represents the column. For a vector there are no two dimensions and only one dimension can be called. For example, `vec1[3]` would yield `3`. Try it. 

Columns can also be called by the `$` sign, but only if a name has been assigned. With dataframes assigning names happens automatically. 

A useful function to inspect the structure of a dataframe is `str()`. Try running it. 

```{r}
str(dat3)
```

Inspecting the structure of your data is vital, as you probably have imported your data from some other source. If we, at a later stage, start analyzing our data without the correct measurement level, we may run into problems. One problem that often occurs is that categorical variables (factors in `R`) are not coded as such (or the reverse). 

---

9. **Imagine that the first variable `V1` in our dataframe `dat3` is not coded correctly, but actually represents grouping information about cities. Convert the variable to a factor and add the labels Utrecht, New York, London, Singapore, Rome and Cape Town.** 
```{r}
dat3$V1 <- factor(dat3$V1,
                  labels = c("Utrecht", "New York", "London", "Singapore", "Rome", "Cape Town"))
dat3
```

Rerunning the `str` function confirms that we now have converted the data in the first column to factors:
```{r}
str(dat3)
```
Note that this conversion was rather lazy, and that the `factor` function offers more controls if needed. 

---

10. **Open the dataset `boys.RData`. You need to download this file and put it in the project folder.**

If you have organized your code in a project the `boys.RData` file will be visible in the Files-panel, where you can just click on it and confirm, that you want to load the data file into your global environment. However, the preferred way is (almost) always to do this using code: 
```{r}
load("boys.RData")
```

Yet another option is to double-click the `boys.RData` file on your machine, eg. in the Windows Explorer (right-click and `open with RStudio` if it does not open by default in `RStudio`, but in `R`). 

---

11. **Most packages have datasets included. Since we have not learned to load packages yet, you are presented with such a data set in a workspace. Open the `boys` dataset (it is from package `mice`, by the way) by typing `boys` in the console, and subsequently by using the function `View()`. ** Tip: To understand the numbers you are looking at look at the documentation by typing `?mice::boys`

The output is not displayed here as it is simply too large. 

Using `View()` is well suited for inspecting datasets that are large (but bot huge). `View()` opens the dataset in a spreadsheet-like window (like Excel), but you cannot edit the content. 

---

12. **Find out the dimensions of the `boys` data set and inspect the first and final 6 cases in the data set. **

To do it numerically, find out what the dimensions of the boys dataset are. 
```{r}
dim(boys)
```
There are 748 cases on 9 variables. To select the first and last six cases, use
```{r}
boys[c(1:6,743:748), ]
```
or, more efficiently:
```{r}
head(boys, n=6)
tail(boys)
```
The functions `head()` and `tail()` are very useful functions. For example, from looking at both functions we can observe that the data are very likely sorted based on `age`.

---

13. **It seems that the `boys` data are sorted based on `age`. Verify this.** 

To verify if the data are indeed sorted, we can run the following command to test the complement of that statement. Remember that we can always search the help for functions: e.g. we could have searched here for `?sort` and we would quickly have ended up at function `is.unsorted()` which tests whether an object is not sorted 

```{r}
is.unsorted(boys$age)
```
which returns `FALSE`, indicating that boys' age is not unsorted, which is the same as saying that it *is* sorted (double negative). To directly test if it is sorted, we could have used
```{r}
!is.unsorted(boys$age)
```
which tests if data data are not unsorted. In other words the values `TRUE` and `FALSE` under `is.unsorted()` turn into `FALSE` and `TRUE` under `!is.unsorted()`, respectively. 

---

14. **Inspect the `boys` dataset with `str()`. Use one or more functions to find distributional summary information (at least information about the minimum, the maximum, the mean and the median) for all of the variables. Give the standard deviation for `age` and `bmi`. ** Tip: make use of the help (?) and help search (??) functionality in `R`. 

```{r}
str(boys)
summary(boys) # summary info
sd(boys$age)  # standard deviation for age
sd(boys$bmi)  # standard deviation for bmi
sd(boys$bmi, na.rm = TRUE) # standard deviation for bmi (observed cases)
```

Note that `bmi` contains 21 missing values, e.g. by looking at the summary information. Therefor we need to use `na.rm = TRUE` to calculate the standard deviation on the observed cases only.

---

15. **Select all boys that are 20 years or older. How many are there?**

The logical operators (TRUE vs FALSE) are a very powerful tool in `R`. For example, we can just select the rows (respondents) in the data that are older than 20 by putting the logical operator within the row index of the dataset:
```{r}
boys2 <- boys[boys$age >= 20, ]
nrow(boys2)
```

or, alternatively, 
```{r}
boys2.1 <- subset(boys, age >= 20)
nrow(boys2.1)
```

To understand the first construction, we can do this in steps by constructing a logical vector `a`:
```{r}
a <- boys$age >= 20
tail(a, n=100)
sum(a)
boys_20 <- boys[a,]
nrow(boys_20)
```


---

16. **Select all boys that are older than 19, but younger than 19.5. How many are there?**
```{r}
boys3 <- boys[boys$age > 19 & boys$age < 19.5, ]
nrow(boys3)
```

or, alternatively, 
```{r}
boys3.2 <- subset(boys, age > 19 & age < 19.5)
nrow(boys3.2)
```

---

17. **What is the mean age of boys younger than 15 years of age that do not live in region `north`?**

```{r}
mean(boys$age[boys$age < 15 & boys$reg != "north" ], na.rm = TRUE)
```

or, alternatively, 
```{r}
mean(subset(boys, age < 15 & reg != "north")$age, na.rm=TRUE)
```

The mean age is `r sprintf("%.2f", mean(boys$age[boys$age < 15 & boys$reg != "north" ], na.rm = TRUE))` years.

--- 

End of practical B.